---
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    pandoc_args: ['-F', 'zotref', '-F', 'pandoc-citeproc']
title: Programming In Base R
author: Yuhao Lin
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  prompt = TRUE,
  cache = FALSE,
  eval = TRUE
)
```

```{r clear global environment, include = FALSE}
remove(list = ls())
```

```{r loading packages, include = FALSE}
library(tidyverse)
```

# Introduction

This session is going to be slightly different from the previous ones,
whereas in the previous sessions we focused on using `tidyverse` for doing
data analysis, in this session we will focus on about base R to complete
some more programming-oriented tasks.

The rationale of this is, it is firstly always good to learn about the
lower level details of things. These `tidyverse` functions have base R
functions as their building blocks (at least in a conceptual sense if not in
the actual implementation).

This will aid your understanding of the higher level abstractions. For
instance we will see in the next session that `mutate_all` can be thought
of as a `for` loop through all the columns of a data frame. Secondly, as
the next session will be focusing on techniques for writing functions, it
is easier to write functions in base R, because `tidyverse` has a different
style of evaluation. To see this, look at the following example:

* `tidyerse` function

```{r}
dat <- mutate(mtcars, cyl = cyl * 2)
```

* Base R function

```{r}
dat["cyl"] <- dat["cyl"] * 2
```

Notice that when referring to the column "cyl" we need quotes for using a
base R function but not for `mutate` (in R `[` indeed is a function).

There are optional exercises at the end of this session, please at least
read and think about them. We will go through them during the session. If
you have questions about your code you could send your script to Yuhao Lin
on Teams before the session. Please also indicate whether you would be
happy for me to use your code as an example during the session (anonymously
of course).

# Data Types

Two basic types of things/objects: **data** and **functions**

- **Data**: things like 7, "seven", $7.000$, and $\left[ \begin{array}{ccc} 7 & 7 & 7 \\ 7 & 7 & 7\end{array}\right]$
- **Functions**: things like `log`, `+` (takes two arguments), `<` (two), `%%` (two), and `mean` (one)

  a function is like a machine which turns input objects, or **arguments**,
  into an output object, or a **return value**(a return value is to be thought
  of as something you can assign to a variable. So if a function prints
  something in the console that you cannot assign it is not a return value
  (one of the main reason that we write user-defined functions is that we
  want to organize our code so we don't have to code out our solutions every
  time).

  It might also do other things like deleting an existing variable.

```{r}
rm(a)
```

Anything else other than outputting a return value that a function does
  is called a side effect.

Note that in R on the outset a function almost never modify an object in
  place but instead would return something, regardless of what the function
  name might imply. Actually there is a function called `modify` in `purrr`
  that returns an object that has the same type as its input.

Basic data types in R:

- **Booleans**:  `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero)
- **Characters**: letters or symbols **strings**: sequences of characters
- **Floating point numbers**: any number with a decimal place, because of
  how float calculations work, rounding is a bit dodgy so watch out’

  (a bit more detail: In R this type is often referred to as `double` or
  `numeric`. Because computers cannot really store a number with arbitrary
  decimal places so a floating point numbers have certain precision.
  `double` refers to "double precision" floating point numbers (there is no
  single precision floating point numbers in R))

- **Missing or ill-defined values**: `NA`, `NaN`, etc.

## Operators

- **Unary**: takes just one argument. E.g., `-` for arithmetic negation,
  `!` for Boolean negation
- **Binary**: take two arguments. E.g., `+`, `-`, `*`, and `/` . Also, `%%`
  (for mod), `%/%` (for integer division) and `^`

```{r}
-7
7 + 5
7 - 5
```

```{r}
7 * 5
7^5
7 / 5
7 %% 5
7 %/% 5
```

## Comparison operators

These are also binary operators; they take two objects, and give back a Boolean

```{r}
7 > 5
7 < 5
7 >= 7
```


```{r}
7 <= 5
7 == 5
7 != 5
```

Warning: `==` is a comparison operator, `=` is not!

## Logical operators

These basic ones are `&` (and) and `|` (or)

```{r}
(5 > 7) & (6 * 7 == 42)
(5 > 7) | (6 * 7 == 42)
(5 > 7) | (6 * 7 == 42) & (0 != 0)
(5 > 7) | (6 * 7 == 42) & (0 != 0) | (9 - 8 >= 0)
```

## More types

- The `typeof()` function returns the data type
- `is.foo()` functions return Booleans for whether the argument is of type *foo*
- `as.foo()` (tries to) "cast" its argument to type *foo*, to translate it sensibly into such a value
- `foo` is just a meaningless placeholder word used commonly in programming
  (also there is `bar`).

```{r}
typeof(7)
is.numeric(7)
is.na(7)
is.na(7 / 0)
7 / 0
is.na(0 / 0)
0 / 0
```


```{r}
is.character(7)
is.character("7")
is.character("seven")
is.na("seven")
```


```{r}
as.character(5 / 6)
as.numeric(as.character(5 / 6))
6 * as.numeric(as.character(5 / 6))
5 / 6 == as.numeric(as.character(5 / 6))
```

# Data Structures

## First data structure: vectors

- A **data structure** is a grouping of related data values into an object
- A **vector** is a sequence of values, all of the same type

Note that although we imply here the distinction between a "value" (an
  element) and a vector (a group) to aid understanding, a "value" such as
  in the example above `as.character(5 / 6)` is really a single element
  vector, which means all of the functions mentioned above actually all
  apply to vectors.


```{r}
x <- c(7, 8, 10, 45)
x
is.vector(x)
```

- The `c()` function (`c` meaning "combine") returns a vector containing
  all its arguments in specified order

- `1:5` is shorthand for `c(1,2,3,4,5)`, and so on `x[1]` would be the
  first element, `x[4]` the fourth element, and `x[-4]` is a vector
  containing *all but* the fourth element


`vector(length=n)` returns an empty vector of length *n*; helpful for filling things up later

```{r}
weekly.hours <- vector(length = 5)
weekly.hours
weekly.hours[5] <- 8
weekly.hours
```

### Vector arithmetic

Arithmetic operator apply to vectors in a "element-wise" fashion

```{r}
x <- c(7, 8, 10, 45)
y <- c(-7, -8, -10, -45)
x + y
x * y
```

### Recycling

**Recycling** repeat elements in shorter vector when combined with a longer
one. Note that to be able to properly repeat elements, the longer vector
needs to have length divisiable by the shorter one.

```{r}
x + c(-7, -8)
x^c(1, 0, -1, 0.5)
x + c(1, 2, 3)
```

Single numbers are vectors of length 1 for purposes of recycling:

```{r}
2 * x
```

---

Can do elementwise comparisons with vectors:

```{r}
x > 9
```

Logical operators also work elementwise:

```{r}
(x > 9) & (x < 20)
```


### Indexing vectors

Often we want to extract elements from a vector, indexing allows us to pick
elements from a vector by knowing where they are (or their labels, which we
will get to) but without needing to know what they are.

Vector of indices:

```{r}
x[c(2, 4)]
```

Vector of negative indices:

```{r}
x[c(-1, -3)]
```

Boolean vector:

Similarly we can use a boolean vector (containing `TRUE` and `FALSE`) to
indicate the positions.

```{r}
x > 9
x[x > 9]
y[x > 9]
```

Pay close attention to how this indexing works. `x > 9` gives us a logical
vector which is then used by `[]` to do the indexing - this is a two-stage
process.

If we use

```{r}
x[c(FALSE, FALSE, TRUE, TRUE)]
```

It would give the same result.

`which()` gives the elements of a Boolean vector that are `TRUE`:

```{r}
places <- which(x > 9)
places
y[places]
```

We can give names to elements of vectors, and index vectors accordingly

```{r}
names(x) <- c("v1", "v2", "v3", "f")
names(x)
x[c("f", "v1")]
```

Note: here R is printing the labels, these are not additional elements of `x`


`names()` returns another vector (of characters):

```{r}
names(y) <- names(x)
sort(names(x))
which(names(x) == "fred")
```

## List

A **list** is sequence of values, but not necessarily all of the same type.
It is mostly to store data with complex (e.g. nested structures) such as
statistical model output.

```{r}
my.list <- list("exponential", 7, FALSE)
my.list
```

Most of what you can do with vectors you can also do with lists

### Accessing pieces of lists

- `[ ]` returns a list.
- `[[ ]]` drops names and structures (returns an unnamed vector)

```{r}
my.list[2]
my.list[[2]]
my.list[[2]]^2
```

Here we emphasize the difference between **list** and **vector**, two
  seemingly similar data structures, because although they appear similar
  the functions we can use on them are different.

Actually most of the commonly used based R functions are only for vectors
  (this is not the case in `tidyverse` (with many functions operating on
  dataframes) which we will see in the next session).

```{r}
mean(c(1, 2, 3))
mean(list(1, 2, 3))
```

### Expanding and contracting lists

Add to lists with `c()` (also works with vectors):

```{r}
my.list <- c(my.list, 9)
my.list
```

Pluck out all but one piece of a list (also works with vectors):

```{r}
my.list[-2]
```

### Names in lists

We can name some or all of the elements of a list:

```{r}
names(my.list) <- c("family", "mean", "is.symmetric")
my.list
my.list[["family"]]
my.list["family"]
```

---

Lists have a special short cut way of using names, with `$`:

```{r}
my.list[["family"]]
my.list$family
```

Adding named elements:

```{r}
my.list$was.estimated <- FALSE
my.list[["last.updated"]] <- "2015-09-01"
```

Note that we are assigning to a non-existent column - that column will be
created upon assignment.

By now you have probably realized that lists are somewhat similar to data
frames - they can have many elements of different types which can be
accessed both with the dollar sign. Internally a data frame column actually is a list.

The key difference between them is a trivial but important one. Data frames
store **tabular** data, in that they are rectangular tables (think of how
each column in a data frame must have the same length) whereas lists can
have elements of different lengths.

## Data frames

We can also access the elements (columns) of a data frame also with `[]`
and `[[]]`. Theoretically, the elments could also be rows or even
individual cell values, but R takes a column (variable) centered approach.

The way of indexing and subsetting dataframes in base R is, as some
  people believe, not entirely intuitive or consistent. That's why we often
  use `tibble` from `tidyverse`, which is essentially an enhanced version
  of data frames but are compatible with all the data frame operations but
  might behave slightly differently in some cases. There is also similarly,
  `data.table` from the `data.table` package, which has a focus on
  performance (so if you have a large data set).

```{r}
mtcars[1]
# returns a data frame
```

```{r}
mtcars_tbl <- as_tibble(mtcars)
mtcars_tbl[1]
# But if the dataframe is a `tibble` it returns a dataframe,
```

```{r}
mtcars[[1]]
# returns a vector
```

Note that we emphasize the difference here because similarly dataframes,
lists and vectors accept different functions.

It should be apparent that we can do `mtcars[c(1, 2)]` but not
`mtcars[[c(1, 2)]]` because a vector is one dimensional. The latter is
actually allowed in R but does not return columns. Experiment with this if
you're interested.

```{r}
mtcars$mpg
# returns a vector
```

We might also need to subset rows of a dataframe. We use this notation,

```{r}
mtcars[c(1, 2), c(2, 3)]
# which means rows 1 and 2, of columns 2 and 3.
```

```{r}
mtcars[c(1, 2), ]
# means rows 1 and 2 of all the column.
```

Similarly we can use this notation with columns

```{r}
mtcars[, c(1, 2)]
# returns a data frame
```

```{r}
mtcars[, 1]
# returns a vector.
```

Data frames can have names (the column names) and they can also be used for
subsetting.

In addition, logical vectors can be used for subsetting as well,  which is
similar to how it is used with vectors. Try these out yourself.


# Control flow

Summary of the control flow tools in R:

- `if()`, `else if()`, `else`: standard conditionals
- `ifelse()`: conditional function that vectorizes nicely

## `if()` and `else`

Use `if()` and `else` to decide whether to evaluate one block of code or another, depending on a condition

```{r}
x <- 0.5

if (x >= 0) {
  x
} else {
  -x
}
```

- Condition in `if()` needs to give one `TRUE` or `FALSE` value
- Note that the `else` statement is optional
- Single line actions don't need braces, i.e., could shorten above to `if (x >= 0) x else -x`

## `else if`

We can use `elseif()` arbitrarily many times following an `if()` statement

```{r}
x <- -2

if (x^2 < 1) {
  x^2
} else if (x >= 1) {
  2 * x - 1
} else {
  -2 * x - 1
}
```

- Each `elseif()` only gets considered if the conditions above it were not `TRUE`
- The `else` statement gets evaluated if none of the above conditions were `TRUE`
- Note again that the `else` statement is optional


## `for` loops

A `for()` loop increments a **counter** variable along a vector. It repeatedly runs a code block, called the  **body** of the loop, with the counter set at its current value, until it runs through the vector

```{r}
log.vec <- c()
for (i in 1:10) {
  log.vec[i] <- log(i)
}
log.vec
```

Here `i` is the counter and the vector we are iterating over is `1:n`. The
body is the code in between the braces. Note that in general the braces are
only used to combine multiple lines of code. So if the body of `for` is
only one line we can use `for(i in 1:10) log.vec[i] = log(i)`.

Notice how we need to "initiate" a empty variable `log.vec` before the
loop. Try running the code with this initialization, what happens?

Let's look another relevant example demonstrating the idea of initialization.

```{r}
sum <- 0
for (i in 1:100) {
  sum <- sum + i
}
print(sum)
```

### Breaking from the loop

We can **break** out of a `for()` loop early (before the counter has been iterated over the whole vector), using `break`

```{r}
n <- 10
log.vec <- vector(length = n, mode = "numeric")
for (i in 1:n) {
  log.vec[i] <- log(i)
  if (log(i) > 2) {
    cat("I'm outta here. I don't like numbers bigger than 2\n")
    break()
  }
}
log.vec
```

There is also `next()`, which allows you to skip to the next iteration.

```{r}
n <- 10
for (i in 1:n) {
  if (i == 2) {
    next()
  }
  print(i)
}
```

Finally notice that in a for loop R does not automatically print to the
console when a command is executed.

```{r}
for (i in 1:10) {
  i * 10
}
```

We need to use `print`. Why? Because in R `for` is actually also a
function. So the results of your computation stays within the function
environment unless you return something, but `for` restricts the return
value to be `NULL` (more on this the next session).

```{r}
for (i in 1:10) {
  print(i * 10)
}
```

### Variations on standard `for()` loops

Many different variations on standard `for()` are possible. Two common ones:

- Non-numeric counters: counter variable always gets iterated over a vector, but it doesn't have to be numeric
- Nested loops: body of the `for()` loop can contain another `for()` loop (or several others)

```{r}
for (str in c("Prof", "Ryan", "Tibs")) {
  print(paste(str, "declined to comment"))
}
```

More generally, any data structure that yo can loop through is called an
iterable. Vectors, lists and data frames (and its variants like `tibble`)
are all iterables. Be careful with what your counter represents in
different data structures (experiment with this yourself!).

```{r}
for (i in 1:4) {
  for (j in 1:i^2) {
    print(paste(j, ""))
  }
  print("ABC")
}
```

How many times `print(paste(j,""))` is executed in the second loop? How do
you calculate this?


# Exercises

## Exercise 1

Translate the following code using `tidyverse` functions to code using base
R functions with the data set from the **Descriptives** session

```{r, eval = F}

dat1 <- dat %>%
  select(
    dem.age,
    dem.alevels,
    dem.employment,
    bmi.cur
  ) %>%
  mutate(
    dem.age_mean_by_sex = mean(dem.age, na.rm = T),
    dem.age_sd_by_sex = sd(dem.age, na.rm = T)
  ) %>%
  rename(
    "Mean" = dem.age_mean_by_sex,
    "SD" = dem.age_mean_by_sex
  ) %>%
  filter(bmi.cur > 20)
```

As the next two exercises mainly concerns writing loops, we need to look a
bit at a simple way to debug which will really help you identity sources of
errors if you make a mistake in your code when writing those loops.

Suppose now we want to write a for loop to print out all the elements in
`x` converted to `numeric`.

```{r}
x <- c("1", "b", "3")
for (i in 1:3) {
  print(as.numeric(x[i]))
}
```

So we get a warning and an `NA`. Now in real life the content of `x` might
not be completely transparent to you, maybe because it might have too many
elements. So we would need a way to investigate what happens within the
loop loop. To do this we use the `print` function.


```{r}
x <- c("1", "b", "3")
for (i in 1:3) {
  print(x[i])
  print(as.numeric(x[i]))
}
```

Now we see "b" cannot be converted to numeric.


## Exercise 2

Use a for loop to generate a Fibonacci sequence up to the 8th term.
The Fibonacci Sequence is the series of numbers:

$0, 1, 1, 2, 3, 5, 8, 13, 21, 34, ...$

The first two numbers are $0$ and $1$. Any number after the first two is
found by adding up the two numbers before it.

**Note: do not use recursion (defining a function that calls itself) or the
closed-form formula for the sequence if you come across them.**

**Hint: the first few lines can be:**

```{r}
i1 <- 0
i2 <- 1
for (i in 1:8) {}
```

```{r, echo = FALSE, eval = F}
i1 <- 0
i2 <- 1
for (i in 1:8) {
  if (i == 1) {
    num <- i1
  } else if (i == 2) {
    num <- i2
  } else {
    num <- i1 + i2
    i1 <- i2
    i2 <- num
  }
  print(num)
}
```

We are going to use another variant of iteration, a `while` loop in our next exercise.

A `while` loop works as follows.

```{r}
i <- 0
while (i < 10) {
  print(i)
  i <- i + 1
}
```

Notice in each iteration we first test if `i < 10`,  print `i` and then add
1 to `i`. At the beginning of the iteration if we realize now `i` is
already larger than 10 we stop. We use a for loop when we know at each
iteration we should end so we can specify `for (i in start:end)`, whereas
we use a `while` loop when we don't know at each iteration we will end.

A typical use case in statistics is when we want to find the values of our
parameters which will maximize the maximum likelihood (MLE) function, this
would be the parameter value at which the derivative of the MLE function is
zero. But for models like logistic regression, there is not values that
would make it exactly zero, and we have to try out parameter values to find
one that makes the derivative closest to zero. We cannot try out all
possible parameter values because there are infinitely many, instead we say
if it is close enough we stop So we would use something like
`while(error > 0.01)` to implement this.

## Exercise 3

Use a `while` loop to implement the following search algorithm and apply it
to a vector.

The following search algorithm determines if a value is in a **sorted** vector:

1. Find the middle element of the vector (e.g. for c(1, 2, 3, 5, 7) this
   would be 3. If the vector length is an even number use the smaller one
   of the two middle numbers)

2. If the element we are searching for is smaller than this middle number
   we restrict the search range to everything smaller than the middle number.

   For example, if we are searching for 2 and 2 is smaller than the middle number 3 in
   `c(1, 2, 3, 5, 7)` we restrict our search range to `c(1, 2)`.

   If the element is larger than the middle number restrict the range to
   anything larger than the middle number.

3. Repeat step 1 and 2 in the restricted range. (e.g. the middle number of
   `c(1, 2)` is 1 and 2 is smaller than 1 so we restrict the range to `c(2)`).

4. If the element is equal to the middle number, print "True". Otherwise if
   the search range has only one number left and is not equal to the
   element we are searching for, print "False". (e.g. 2 is equal to 2 in
   `c(2)` so we print `True`).

**Hint: the code might look like this**

```{r, eval = F}
# Given an ordered sequence
x <- c(1, 2, 3, 4, 5, 6, 9, 10)

# Initiate variables to indicate our search range
start <- 1
end <- length(x)

# We try to find the element 11 (not in the sequence)
while (x[i] != 11 & "some condition") {
  i <- (start + end) %/% 2
}

# This is the last line of your code
if (x[i] == 11) print("TRUE") else print("FALSE")
```

```{r, eval = F}
# Make sure it works for an element that's in the sequence
while (x[i] != 9 & "some condition") {}
```

```{r, eval = F, echo = F}
x <- c(1, 2, 3, 4, 5, 6, 9, 10)
start <- 1
end <- length(x)
while (x[i] != 9 & start != end) {
  i <- (start + end) %/% 2
  if (x[i] < 9) {
    start <- i
  } else {
    end <- i
  }
}
if (x[i] == 9) print("TRUE") else print("FALSE")
```
